Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-05 article-models-stage2.tex
written rule for weakening followed by substitution

>>>>

\section{Stage three: Beyond}
We also pose the following questions which, I'm afraid, are beyond the scope
of the proposed research program.

\begin{enumerate}
\item In this document we have developed a custom type theory without basic
constructors. It is not a question whether someone will want to propose
a slightly different type theory without basic constructors, someone will. So
it would be nice if there were a type theory of type theories, formalizing what
a type theory actually is. Then one could define his or her new type theory
as an object of that theory. Moreover, it could then be possible to write a
program that automatically generates a theorem prover for user defined type
theories in this language, so that a person who has come up with a new type
theory (such as Guillaume who came up with his type theory of higher groupoids)
can immediately start giving formally verified proofs in his new type theory.
One could also envision that there will be morphisms in this type theory which
automatically transform a formalized development of a theory from one type
theory to another, so that not every simple thing has to be done over and over
again. So the question is, could something like that actually be possible? We
may actually use our type theory before type constructors for that. Or, if we
manage to get hold on the internal logic of a model
(which should be a type theory such that the model in question is its initial
model?)
for each object of our type theory of models of type theory, we might actually
already have a `type theory of type theories'.
\end{enumerate}

### 2014-03-05 article-models-tt.tex
written rule for weakening followed by substitution

@@

\subsubsection{Extension acts as the identity on families of families}
The family of which $\tfext{A}$ is a term, is unfolded as
\begin{equation*}
\unfold{\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
\end{equation*}
Therefore, there is a term
\begin{equation*}
\jhom
  {{\Gamma}{A}}
  {\mfam{\mfam{A}}}
  {\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
\end{equation*}
By the rule asserting that families of families are families of extensions, we
have $\mfam{\mfam{A}}\jdeq\jcomp{}{\tfext{A}}{\mfam{A}}$. By the rule that
families over weakenings are weakenings of families, we have the judgmental
equalities
\begin{equation*}
\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}
\jdeq 
  \unfold{\jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}}
\jdeq 
  \jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}
\end{equation*}
Therefore, we can compare the term $\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}$
to the identity function $\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]$. 
The compatibility
rule for the action of extension on families of families asserts that
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
      {{\Gamma}{A}}
      {\jcomp{}{\tfext{A}}{\mfam{A}}}
      {\jcomp{}{\tfext{A}}{\mfam{A}}}
      {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
      {\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]}
    }
\end{equation*}
\emph{(Note: this rule might be a consequence of an explanation of what $\mfam{{A}{P}}$
is in general, but at the moment I don't see how to do this)}

>>>>

Now let $\jfam{{\Gamma}{A}}{P}$ and $\jfam{{\Gamma}{A}}{Q}$ be families. A
morphism from $P$ to $Q$ over the identity function $\idfunc[A]$ in context
$\Gamma$ is the same thing as a morphism from $P$ to $Q$ in context
$\ctxext{\Gamma}{A}$:

\begin{lem}\label{hom-over-id-is-hom}
We have the following valid inference rules:
\begin{align*}
& \inference
  { \jfam{{\Gamma}{A}}{P}
    \jfam{{\Gamma}{A}}{Q}
    \jfhom{\Gamma}{A}{A}{\idfunc[A]}{P}{Q}{f}
    }
  { \jhom{{\Gamma}{A}}{P}{Q}{f}
    }
  \\
& \inference
  { \jfam{{\Gamma}{A}}{P}
    \jfam{{\Gamma}{A}}{Q}
    \jhom{{\Gamma}{A}}{P}{Q}{f}
    }
  { \jfhom{\Gamma}{A}{A}{\idfunc[A]}{P}{Q}{f}
    }
\end{align*}
\end{lem}

\begin{proof}
If we unfold the judgments $\jhom{{\Gamma}{A}}{P}{Q}{f}$ and
$\jfhom{\Gamma}{A}{A}{\idfunc[A]}{P}{Q}{f}$, we get the judgments
\begin{align*}
& \unfold{\jhom{{\Gamma}{A}}{P}{Q}{f}}
  \\
& \unfold{\jfhom{\Gamma}{A}{A}{\idfunc[A]}{P}{Q}{f}}
\end{align*}
respectively. Therefore, we only need to verify that
$\ctxwk{P}{\subst{\idfunc[A]}{\ctxwk{A}{Q}}}\jdeq\ctxwk{P}{Q}$, which is indeed
the case by \autoref{idfunc-wk-defn}.
\end{proof}

@@

\subsubsection{Terms of families over families are terms of families over extensions}
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq
      {{{{\Gamma}{A}}{\mfam{A}}}{\mfam{\mfam{A}}}}
      {\mtm{\mfam{A}}}
      {\jcomp{}{\tfext{A}}{\mtm{A}}}
      }
\end{equation*}

\subsubsection{Extension acts as the identity on families of families}
The family of which $\tfext{A}$ is a term, is unfolded as
\begin{equation*}
\unfold{\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
\end{equation*}
Because we have the judgmental equality $\ctxext{\Gamma}{{A}{\mfam{A}}}
\jdeq \ctxext{{\Gamma}{A}}{\mfam{A}}$, there is a term
\begin{equation*}
\jhom
  {{{\Gamma}{A}}{\mfam{A}}}
  {\mfam{\mfam{A}}}
  {\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
\end{equation*}
By the rule asserting that families of families are families of extensions, we
have $\mfam{\mfam{A}}\jdeq\jcomp{}{\tfext{A}}{\mfam{A}}$. By the rule that
families over weakenings are weakenings of families, we have the judgmental
equalities
\begin{equation*}
\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}
\jdeq 
  \unfold{\jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}}
\jdeq 
  \jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}
\end{equation*}
Therefore, we can compare the term $\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}$
to the identity function $\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]$. 
The compatibility
rule for the action of extension on families of families asserts that
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
      {{{\Gamma}{A}}{\mfam{A}}}
      {\jcomp{}{\tfext{A}}{\mfam{A}}}
      {\jcomp{}{\tfext{A}}{\mfam{A}}}
      {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
      {\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]}
    }
\end{equation*}
\emph{(Note: this rule might be a consequence of an explanation of what $\mfam{{A}{P}}$
is in general, but at the moment I don't see how to do this)}

\subsubsection{Extension acts as the identity on terms of families of families}
We now investigate the nature of the morphism.
\begin{equation*}
\jfhom
  {{{\Gamma}{A}}{\mfam{A}}}
  {\mfam{\mfam{A}}}
  {\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
  {\mtm{\mfam{A}}}
  {\subst{\tfext{A}}{\mtm{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mtm{\tfext{A}}^{\ctxext{\Gamma}{A}}}
\end{equation*}
Note that we have the judgmental equality 
$\mtm{\mfam{A}}\jdeq\jcomp{}{\tfext{A}}{\mtm{A}}$. Likewise, we have the
judgmental equality $\subst{\tfext{A}}{\mtm{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}
\jdeq\jcomp{}{\tfext{A}}{\mtm{A}}$. Thirdly, we have the judgmental equality
$\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}\jdeq\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]$.
We may combine these three facts with \autoref{hom-over-id-is-hom} to see that
\begin{equation*}
\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\jcomp{}{\tfext{A}}}{\mfam{A}}}
  {\jcomp{}{\tfext{A}}{\mtm{A}}}
  {\jcomp{}{\tfext{A}}{\mtm{A}}}
  {\mtm{\tfext{A}}^{\ctxext{\Gamma}{A}}}
\end{equation*}
Now we see that we can require that $\mtm{\tfext{A}}^{\ctxext{\Gamma}{A}}$ is
the identity function on $\jcomp{}{\tfext{A}}{\mtm{A}}$:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
    {{{{\Gamma}{A}}{\mfam{A}}}{\jcomp{}{\tfext{A}}}{\mfam{A}}}
    {\jcomp{}{\tfext{A}}{\mtm{A}}}
    {\jcomp{}{\tfext{A}}{\mtm{A}}}
    {\mtm{\tfext{A}}^{\ctxext{\Gamma}{A}}}
    {\idfunc[\jcomp{}{\tfext{A}}{\mtm{A}}]}
    }
\end{equation*}

\subsubsection{Weakening followed by substitution is the identity}
The weakening morphism $\tfwk{A}^0$ gives a morphism
\begin{equation*}
\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\tfwk{A}^0}}
\end{equation*}
which can be composed with the morphism $\tfsubst{A}^0$. We require that this
composition is the identity function on $\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
      {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
      {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
      {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
      {\ctxwk{\mtm{A}}{\tfwk{A}^0}}
      {\idfunc[\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}]}
    }
\end{equation*}
